---
title: "Regression Analysis of Water Systems"
author: "John Meroth"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## DRINKING WATER VIOLATIONS

The following logistic regression analysis in R estimates the probability of a violation occurring based upon the chosen feature vectors.  For most choices, it is believed that no correlation will be seen (ROC plot a straight line).  In the event a collection of useful features are discovered, the AUC of the ROC plot will increase and the associated model may prove useful in highlighting at-risk water systems.

Notes:  
Sparse data exists prior to 2008, therefore we use data from 2008 and later.  
Only data from currently "Active" water systems are considered.  
Although some systems report violations quarterly, some report only annually.

## ORGANIZING TEST DATA

The EPA's SDWIS tables WATER_SYSTEM, VIOLATION, and WATER_SYSTEM_FACILITY can be used to generate a set of labelled test data.  Each record, identified by "PWSID" represents a water system in the United States.  

### Setup the environment
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(dummies)
library(readr)
library(tidyverse)
library(pROC)
library(lubridate)
library(effects)
```

### Import data from pwsid.csv
This file was created by querying the WATER_SYSTEM table for all active water systems.  
`SQL code:
    SELECT PWSID
    FROM WATER_SYSTEM
    WHERE PWS_ACTIVITY_CODE = 'A'`
```{r}
getwd()
setwd('C:/Users/jmeroth/WATER/derived')
df <- read_csv("pwsid.csv")
summary(df)
```

### Import data from vHealth.csv
This file was created by filtering the VIOLATION table for health-based violations.  
`SQL code:
    SELECT
    PWSID
    , cast(VIOLATION_ID AS TEXT) VIOLATION_ID
    , COMPL_PER_BEGIN_DATE
    FROM VIOLATION
    WHERE
    IS_HEALTH_BASED_IND = "Y"`
```{r}
setwd('C:/Users/jmeroth/WATER/derived')
vH <- read_csv("vHealth.csv", col_names = TRUE, cols(VIOLATION_ID = "c"))
summary(vH)
```

Convert character field to date:
```{r}
vH <- vH %>% mutate(COMPL_PER_BEGIN_DATE = as.Date(COMPL_PER_BEGIN_DATE, format = "%d-%B-%y"))
str(vH)

```

Create label field "H_viol_2017" from health violation data:
```{r}
vH <- vH %>% 
  filter(year(vH$COMPL_PER_BEGIN_DATE) == 2017) %>%
  group_by(PWSID) %>%
  summarise(H_viol_2017 = 1)
head(vH)
```


Merge data into df forming the test data set:
```{r}
df <- merge(df, vH, by.x="PWSID", by.y="PWSID", all.x=TRUE)
df[is.na(df)] <- 0
summary(df)
```

## FEATURE ENGINEERING

"Features" are characteristics or measurments of each sample in our data set.  A good feature will be somehow related to the chances of a system registering a violation.  The exact relationship to a violation need not be know for the feature to be useful.

### TREATED_SOURCE

Add a binary predictor variable, "TREATED_SOURCE" (1 if treatment facility exists).
The file treated_source.csv was created using this SQL code:  
    with active_facilities AS (
    select ws.PWSID
    from WATER_SYSTEM2 ws left join WATER_SYSTEM_FACILITY_3 fa
    on ws.PWSID = fa.PWSID
    where ws.PWS_ACTIVITY_CODE = 'A'
    and fa.IS_SOURCE_TREATED_IND = 'Y'
    and fa.FACILITY_ACTIVITY_CODE = 'A')
    Select DISTINCT PWSID from active_facilities

```{r}
setwd("C:/Users/jmeroth/water/derived")
treated_source <- read_csv("treated_source.csv")
treated_source$TREATED_SOURCE = 1
df <- merge(df, treated_source, by.x="PWSID", by.y="PWSID"
		, all.x=TRUE)
df[is.na(df)] <- 0
```


### VIOLATION_CATEGORY_CODE

Add a group of variables indicating the occurrance of a non-health violation in the prior year (2016).
The file violation_category_code.csv was created using this SQL code:  
    SELECT
    PWSID, VIOLATION_CATEGORY_CODE, substr(COMPL_PER_BEGIN_DATE, -2) AS vYear
    FROM VIOLATION2
    WHERE
    IS_HEALTH_BASED_IND = "N"
    and PWS_ACTIVITY_CODE = 'A'
    and vYear = '16'
    GROUP by PWSID, VIOLATION_CATEGORY_CODE, vYear  
The resulting records were pivoted by PWSID on VIOLATION_CATEGORY_CODE.

```{r}
setwd("C:/Users/jmeroth/water/derived")
violation_category_code <- read_csv("violation_category_code.csv")
df <- merge(df, violation_category_code, by.x="PWSID", by.y="PWSID"
		, all.x=TRUE)
df[is.na(df)] <- 0
```


### PRIMARY_SOURCE_CODE

The file primary_source_code.csv was created using this SQL code:  
    SELECT PWSID, PRIMARY_SOURCE_CODE
    FROM WATER_SYSTEM
    WHERE PWS_ACTIVITY_CODE = 'A'
    
Add a categorical variable "PRIMARY_SOURCE_CODE":
```{r}
setwd("C:/Users/jmeroth/water/derived")
psc <- read_csv("primary_source_code.csv")
# Merge the new feature vector:
df <- merge(df, psc, by.x="PWSID", by.y="PWSID", all.x=TRUE)
# Convert empty cells to zero:
df[is.na(df)] <- 0

```


### PWS_TYPE_CODE

The file pws_type_code.csv was created using this SQL code:  
    *SELECT PWSID, PWS_TYPE_CODE
    FROM WATER_SYSTEM
    WHERE PWS_ACTIVITY_CODE = 'A'*
    
Add a categorical variable "PWS_TYPE_CODE":
```{r}
setwd("C:/Users/jmeroth/water/derived")
ptc <- read_csv("pws_type_code.csv")
# Merge the new feature vector:
df <- merge(df, ptc, by.x="PWSID", by.y="PWSID", all.x=TRUE)
# Convert empty cells to zero:
df[is.na(df)] <- 0

```


### PRIOR YEAR TOTAL RAINFALL

The file wsAnnualRain.csv was created using the Jupyter Notebook: match_water_weather.ipynb (from Mia Polansky, June 20, 2019)
This file contains the geolocated watersystems along with the annual rainfall at the nearest weather station.
    
Add a normalized, continuous variable "r2016n":
```{r}
# Read in the rainfall data:
setwd("C:/Users/jmeroth/WATER/Weather")
rain <- read_csv("wsAnnualRain.csv")

# Normalize the rainfall data:
rain$X1 <- NULL
colnames(rain)[colnames(rain)=="0"] <- "r2016"
rmin <- min(rain$r2016)
rmax <- max(rain$r2016)
rscale = rmax-rmin
rain["r2016n"] <- (rain$r2016-rmin)/rscale  - 1.0
rain$r2016 <- NULL
```

```{r}
# Merge the new feature vector:
df <- merge(df, rain, by.x="PWSID", by.y="PWSID", all.x=TRUE)
# Convert empty cells to zero:
df[is.na(df)] <- 0

```


## MODELLING


### Run the Logistic Regression Model:


```{r}
logit <- glm(H_viol_2017 ~ TREATED_SOURCE + PRIMARY_SOURCE_CODE + PWS_TYPE_CODE + MON + MR + Other + RPT + r2016n
 , data = df, family = "binomial")
summary(logit)

```

Result of regression using the "effects" library:
```{r}

plot(effects::allEffects(logit), axes=list(grid=FALSE))
```

### Create Test Data
```{r}
test_data <- df
test_data$PWSID <- NULL
test_data$H_viol_2017 <- NULL
```



Calculate probabilities P, add to data frame.
```{r}
df["P"] <- predict(logit, test_data, type="response")
```


## ANALYSIS

### Analysis of Means
```{r}
df %>% group_by(H_viol_2017) %>% summarise(mean = mean(P))
```

### Pivot tables
```{r}
xtabs(~H_viol_2017 + PRIMARY_SOURCE_CODE, data = df)
```

```{r}
xtabs(~H_viol_2017 + PWS_TYPE_CODE, data = df)
```

### Probability Distribution
```{r, echo=FALSE}
hist(df$P)
```


### Confusion Marix with threshold set as P = 0.3
```{r}
df %>% group_by(P > 0.3) %>% filter(H_viol_2017 == 1) %>% summarise(n())
df %>% group_by(P > 0.3) %>% filter(H_viol_2017 == 0) %>% summarise(n())
```


### ROC
```{r message=FALSE, warning=FALSE}
#Create a square plot
par(pty = "s")
roc(df$H_viol_2017, df$P, plot=TRUE, legacy.axes=TRUE, percent=FALSE, xlab="False Positive Rate", ylab="True Positive Rate", col="#377eb8", lwd=4, print.auc=TRUE)

```


## FUTURE WORK

Two years of rainfall data should be included as well as demographic data.
